{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}

### 2018-01-24 sftrends_manu.tex
minor edits from Thomas and Phil comments

\begin{singlespace}
\title{{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^{1*}$, Thomas W. Jabusch$^2$, Philip R. Trowbridge$^2$, David B. Senn$^2$}}
  \\\\{\textit {\normalsize $^*$Corresponding author: \href{mailto:marcusb@sccwrp.org}{marcusb@sccwrp.org}}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
  \\\\{\textit {\normalsize Current address: Southern California Coastal Water Research Project}}
  \\{\textit {\normalsize 3535 Harbor Blvd, Suite 110, Costa Mesa, CA 92626}}
  \\\\{\textit {\normalsize $^2$San Francisco Estuary Institute}}
	\\{\textit {\normalsize 4911 Central Ave, Richmond, CA 94804}}
  \vspace{1in}
  \\ Version Date:   Fri Oct 27 15:55:09 2017 -0700
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

\section*{Abstract}
\noindent Quantitative descriptions of water quality, physical, and biological characteristics of estuaries are critical for developing an ecological understanding of drivers of change.  Historical trends and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total) from the Delta region of the \acl{sfe} were modeled with an estuarine adaptation of the \ac{wrtds}.  Trend analysis with flow-normalized results demonstrated the potential to interpret different changes relative to observed data that included flow effects, such that several trends with flow-normalized data had changes in magnitude and even reversal of trends relative to the observed. Modelled trends demonstrated that nutrient concentrations were on average higher in the last twenty years relative to the earlier periods of observation, although a consistent decline began in the mid-1990s and early 2000s. We further described mechanisms of change with two case studies that evaluated 1) downstream changes in nitrogen following upgrades at a wastewater treatment plant, and 2) interactions between biological invaders, chlorophyll, and flow in Suisun Bay.  \ac{wrtds} results for ammonium trends showed a distinct signal as a result of upstream \acl{wwtp} upgrades, with specific reductions observed in the winter months during low-flow conditions. Results for Suisun Bay showed that \acl{chla} production in early years was directly stimulated by flow, whereas the relationship with flow in later years was indirect and influenced by grazing pressure. Although these trends and potential causes of change have been described in the literature, results from \ac{wrtds} provided an approach to test alternative hypotheses of spatiotemporal drivers of nutrient dynamics in the Delta.
\acresetall

@@

Nutrient time series of monthly observations from 1976 to 2013 were obtained for ten active sampling stations in the Delta \citep[\cref{fig:delt_map,tab:stdescrp},][]{IEP13}.  Stations were grouped by location in the study area for comparison: \textit{peripheral} Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; \textit{interior} Delta stations D19, D26, D28; and \textit{Suisun} stations D4, D6, and D7.  These stations cover all of the major inflows and outflows to the Delta and were selected for analysis based on the continuity of the period of observation \citep{Jabusch09}. Although many other stations are available for the region, the stations were chosen because they are actively maintained by the regional monitoring program and they capture dominant seasonal and annual modes of nitrogen variability characteristic of the region \citep{Jabusch16}. Time series were complete for all stations except for an approximate ten year gap from 1996-2004 for D19.  Data were minimally processed, with the exception of averaging replicates that occurred on the same day.  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two). Less than 3\% of all observations were below the detection limit (left-censored), although variation was observed between analytes and location.  The ammonium time series had the most censored observations at sites C10 (25.4\% of all observations), MD10 (18.1\%), D28 (17.8\%), D19 (12\%), and D7 (7.9\%).

\ac{wrtds} models require flow data paired with nutrient data. At the peripheral and interior stations, daily flow estimates were matched with the corresponding sample dates for the nutrient data. Daily flow estimates were obtained from the Dayflow software program \citep{IEP16}.  The Sacramento daily flow time series was used to account for flow effects at C3, D19, D26, D28, and MD10, and the San Joaquin time series was used for C10 and P8 based on station proximity to each inflow. Given the complexity of inflows and connectivity of the system, only the inflow estimates from the Sacramento and San Joaquin rivers were used as measures of freshwater influence at each station.  Initial analyses indicated that model fit was not significantly improved with flow estimates from locations closer to each station, nor was model fit improved using lagged times series. Salinity was used as a proxy for flow at sites D4, D6, and D7 where tidal influences were much stronger. Salinity has been used as a tracer of freshwater influence for the application of \ac{wrtds} models in tidal waters \citep{Beck15}.  Models were evaluated using salinity at the interior and peripheral stations but performance was reduced relative to models that used daily flow estimates.

@@

The observed time series for the ten Delta - Suisun Bay stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).  \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.  However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D26) had higher concentrations of ammonium during winter months when assimilation rates are lower \citep{Novick15}.  By location, observed concentrations of \ac{din} for the entire time series were higher on average for the peripheral stations (C3, C10, MD10, P8; mean $\pm$ s.e.: $1.04$$\pm$$0.03$ \mgl) and similar for the interior (D19, D26, D28, $0.43$$\pm$$0.01$) and Suisun Bay stations (D4, D6, D7, $0.44$$\pm$$0.01$).  Average concentrations were highest at P8 ($1.63$$\pm$$0.05$ \mgl) and lowest at C3 ($0.4$$\pm$$0.01$) for \ac{din}, highest at P8 ($0.28$$\pm$$0.02$) and lowest at D28 ($0.05$$\pm$$0.003$) for ammonium, and highest at C10 ($1.4$$\pm$$0.04$) and lowest at C3 ($0.15$$\pm$$0.004$) for nitrite/nitrate. Mean observed concentrations were also higher later in the time series for all analytes.  For example, average \ac{din} across all stations was $0.61$$\pm$$0.01$ \mgl for 1976-1995, compared to $0.7$$\pm$$0.01$ for 1996-2013. Seasonal changes across all years showed that nitrogen concentrations were generally lower in the summer and higher in the winter, although observed patterns were inconsistent between sites.  For example, site MD10 had distinct seasonal spikes for elevated \ac{din} in the winter, whereas other stations had less prominent seasonal maxima (e.g., C3, D7, \cref{fig:obsdat}).

Relative to the observed data, long-term trends between stations for the different nitrogen analytes were apparent from the modelled results (\cref{fig:prddat}).  Although each station varied in the overall concentrations, patterns within the three Delta regions (peripheral, interior, and Suisun) were observed.  Concentrations for all nitrogen analytes were highest in the peripheral stations.  Ammonium concentrations at P8 and C3 were highest and showed a consistent increase over time, followed by a reduction beginning in the early 2000s, whereas ammonium concentrations at C10 and MD10 were low and gradually decreasing throughout the period of record.  By contrast, \ac{din} and nitrite/nitrate concentrations at the peripheral stations showed increases at  P8 and C10 followed by a decline in the early 2000s, whereas concentrations at C3 and MD10 were lower and did not show any noticeable trends.  Trends in the interior stations showed a gradual increase in ammonium followed by a gradual decrease beginning in the early 1990s, particularly for D26.  Trends in \ac{din} and nitrite/nitrate for the interior stations showed a reduction early in the time series, followed by a slight increase beginning in the mid-1980s, and finally a reduction beginning in the late 1990s.  These trends were similar for the Suisun stations, although the reduction in the late 1990s did not occur.  By contrast, ammonium concentrations were low in Suisun but a gradual increase over the period of record was observed.

@@

Water quality conditions in the Delta are dynamic and not easily characterized from observed time series.  Annual aggregations of \ac{wrtds} modelled results and application of formal trend analyses provided insight into the spatial and temporal variation of nitrogen analytes in three distinct regions of the Delta that is not possible with raw observations.  A general conclusion is that nitrogen concentrations have showed a consistent decrease beginning in the mid-1990s and early 2000s, although average concentrations remain above those observed earlier in the period of record.  These results are confirmed visually from \ac{wrtds} (\cref{fig:prddat}) and through significant results from trend tests (\cref{fig:trndcomp1}, \cref{tab:trndsann}).  Although the overall trends suggest a system-wide reduction, considerable differences by location and analyte were characterized by the analysis and are important independent of overall trends.  Nutrient concentrations were highest at the peripheral stations (C3, MD10, P8, and C10) that monitor inflows from the Sacramento, San Joaquin, Cosumnes, and Mokelumne rivers.  The highest concentrations among all nitrogen analytes, ammonium in particular, were observed at P8 as a direct consquence of \ac{wwtp} inputs upstream.  Elevated ammonium concentrations were also observed at C3 as a measure of upstream contributions from the Sacramento Regional County Sanitation District.  By contrast, nitrite/nitrate concentrations were highest at C10 as a measure of contributions from the San Joaquin River to the south that drains a predominantly agricultural watershed.  Although the Sacramento River drains a much larger area, the dominant ammonium signal as compared to nitrite/nitrate at C3 underscores the importance of \ac{wwtp} control for water quality issues downstream.

Differing magnitudes of nitrogen analytes between stations as a function of source type can have an effect on the relationship between flow and nutrients.  Both \citet{Hirsch10} and \citet{Beck15} used \ac{wrtds} results to demonstrate variation between flow and nutrient dynamics depending on pollutant sources.  In particular, a chemodynamic response of nutrients with flow variation is common if nutrients originate primarily from the watershed through diffuse sources \citep{Thompson11,Wan17}.  Increased flow may induce a change in nutrient concentrations, such that reduction may occur with flushing or an increase may occur through mobilization.  By contrast, nutrient loads are relatively chemostatic or invariant with changes in flow if point-sources are the dominant contributor.  These relationships are modelled particularly well with \ac{wrtds}, which can provide a means of hypothesizing unknown sources or verifying trends in response to management actions.  As noted above, C10 at the inflow of the San Joaquin River is dominated by nitrite/nitrate consistent with diffuse, agricultural inputs from the watershed.  A logical expectation is that trends from observed data may vary considerably from trends with modelled results that are flow-normalized.  Accordingly, trend analysis of nitrate/nitrate by year and season showed that percent changes at C10 were typically underestimated with the observed data during the recent period from 1996-2013 (\cref{tab:trndsann,tab:trndsaft}).  This is consistant with an expected effect of flow on raw time series, particularly for chemodynamic behavior at locations that drain highly developed watersheds \citep{Wan17}.

% obs v flow norm trends
Our results demonstrated the potential to interpret different trends if flow effects are not considered, where the difference could vary from a simple change in the magnitude and significance of a trend, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed (e.g., \ac{din} trends for all Suisun stations from 1996-2013, \cref{fig:trndcomp1}). Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes, particularly if trend evaluation is used to assess the effects of nutrients on ecosystem health or the effectiveness of past nutrient management actions. An alternative evaluation of flow in the Delta demonstrated that flow contributions from different sources vary considerably over time at each station \citep{Novick15}.  For example, flow at MD10 represents a changing percentage by season of inputs from the Sacramento, San Joaquin, Cosumnes, Mokelumne rivers, and agricultural returns.   For simplicity, water quality observations in our analyses were matched with large-scale drivers of flow into the Delta where most sites were matched to Sacramento or San Joaquin daily flow estimates.  Given that substantial differences with flow-normalized results were apparent from relatively coarse estimates of flow contributions, more precise differences could be obtained by considering the influence of multiple flow components at each location. Output from the Dayflow software program \citep{IEP16} provides a complete mass balance of flow in the Delta that could be used to develop a more comprehensive description.  Our analysis is the first attempt to model nutrient dynamics related to flow in the entire Delta, such that additional work should focus on improving the characterization of flow signals at each station.

Long-term trends in nutrient and phytoplankton concentations in Suisun Bay have also been the focus of intense study for many years \citep{Cloern83,Lehman92,Dugdale07,Jassby08,Glibert14}.  Although nitrite/nitrate concentrations generally exceed ammonium about five-fold at the Suisun Bay stations, changes in ammonium concentration below 0.072 mg L$^{-1}$ (4 $\mu$mol L$^{-1}$) is a concern given the affect on the uptake of nitrate by phytoplankton \citep{Dugdale07}.  Energetic costs of ammonium are lower than nitrate for phytoplankton growth.  Although algal communities in \ac{sfe} generally utilize ammonium when growth conditions are favorable, seasonal variation in the dominant forms has contributed to occurrence of bloom events in recent years.  In particular, reduction of ammonium in the spring in Suisun Bay below thresholds of  4 $\mu$mol L$^{-1}$ has contributed to uptake of nitrate that stimulates bloom development. Our results demonstrated an overall increase in ammonium from the late 1970s to 2000, with initial flow-normalized values of annual averages estimated as approximately 0.05 mg L$^-1$ for the Suisun stations in 1976 to a maximum ranging from 0.08 mg L$^{-1}$ (station D4) to above 0.1 mg L$^{-1}$ (D6) in 2000 (\cref{fig:prddat}).  Trends from 1996-2005 evaluated in \citep{Jassby08} showed a similar increase in ammonium.  However, a reversal of trends in recent years may also be occuring in Suisun Bay, as model estimates suggest either relatively constant concentrations or even a decrease at some stations beginning around 2000 (e.g., D6, D7, \cref{fig:trndcomp1}).  Combined with the shift towards a dominant spring peak in chloropyll growth (\cref{fig:clmchl}f), changing nitrogen ratios continue to be a concern for the management of production in the upper \ac{sfe}.

@@

The results for Suisun Bay provide additional descriptions of change in production as it relates to flow, grazing, and nitrogen ratios.  In general, clam biomass was associated with a decrease in \ac{chla} concentration, as shown by others \citep{Alpine92,Thompson08}. Our results also suggested that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.  This supports past research that showed a decrease in silica uptake by diatoms following invasion \citep{Cloern96,Kimmerer05}.  Anontrivial portion of the \ac{din} increase could be related to the decrease in a major `sink', i.e., decreased \ac{din} uptake by phytoplankton due to top down grazing pressure from \textit{P. amurensis}.  Flow effects on phytoplankton production have also changed over time. In the absence of benthic grazing prior to invasion, \ac{chla} production was limited at low flow as less nutrients were exported from the Delta, stimulated as flow increases, and reduced at high flow as either nutrients or phytoplankton biomass are exported to the estuary (\cref{fig:clmchl}j). Recent years have shown a decrease in overall chlorophyll, with particularly low concentrations at low flow (high salinity).  As such, \ac{chla} production in early years is directly related to flow, whereas the relationship with flow in later years is indirect as increased flow reduces clam abundance and releases phytoplankton from benthic grazing pressure. These relationships have been suggested by others \citep{Cloern83,Alpine92,Parchaso02,Jassby08}, although the precise mechanisms demonstrated by \ac{wrtds} provide additonal quantitative evidence of factors that drive water quality in the Delta.

@@

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth,page=1]{figs/delt_map.pdf}
\caption{The \acl{sfe} and Delta region with monitoring stations used for analysis. The Delta drains the combined watersheds of the Sacramento and San Joaquin rivers (inset). All data were obtained from the \acl{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}, \cite{IEP13}). See \cref{tab:stdescrp} for station descriptions.}
\label{fig:delt_map}   
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth,page=1]{figs/trndcomp1.pdf}
\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes. Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$). Trends are shown separately for different annual groupings. See \cref{fig:trndcomp2,fig:trndcomp3} for seasonal groupings.}
\label{fig:trndcomp1}   
\end{figure}

@@

\caption[Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County]{Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County.  Wastewater discharge requirements were implemented in 2006 for nitrification/denitrification and tertiary filtration to convert ammonium to nitrate.}\label{fig:stock}
\end{figure}

>>>>

\begin{singlespace}
\title{{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^{1*}$, Thomas W. Jabusch$^2$, Philip R. Trowbridge$^2$, David B. Senn$^2$}}
  \\\\{\textit {\normalsize $^*$Corresponding author: \href{mailto:marcusb@sccwrp.org}{marcusb@sccwrp.org}}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
  \\\\{\textit {\normalsize Current address: Southern California Coastal Water Research Project}}
  \\{\textit {\normalsize 3535 Harbor Blvd, Suite 110, Costa Mesa, CA 92626}}
  \\\\{\textit {\normalsize $^2$San Francisco Estuary Institute}}
	\\{\textit {\normalsize 4911 Central Ave, Richmond, CA 94804}}
  \vspace{1in}
  \\ Version Date:   Fri Oct 27 15:58:09 2017 -0700
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

\section*{Abstract}
\noindent Quantitative descriptions of chemical, physical, and biological characteristics of estuaries are critical for developing an ecological understanding of drivers of change.  Historical trends and relationships between key species of dissolved inorganic nitrogen (ammonium, nitrate/nitrite, total) from the Delta region of the \acl{sfe} were modeled with an estuarine adaptation of the \ac{wrtds}.  
Analysis of flow-normalized data revealed trends that were different from those in the observed time-series. Flow-normalized data exhibited changes in magnitude and even reversal of trends relative to the observed data. Modelled trends demonstrated that nutrient concentrations were on average higher in the last twenty years relative to the earlier periods of observation, although concentrations have been slowly declining since the mid-1990s and early 2000s. We further describe mechanisms of change with two case studies that evaluated 1) downstream changes in nitrogen following upgrades at a wastewater treatment plant, and 2) interactions between biological invaders, chlorophyll, macro-nutrients (nitrogen and silica), and flow in Suisun Bay.  \ac{wrtds} results for ammonium trends showed a distinct signal as a result of upstream \acl{wwtp} upgrades, with specific reductions observed in the winter months during low-flow conditions. Results for Suisun Bay showed that \acl{chla} production in early years was directly stimulated by flow, whereas the relationship with flow in later years was indirect and influenced by grazing pressure. Although these trends and potential causes of change have been described in the literature, results from \ac{wrtds} provided an approach to test alternative hypotheses of spatiotemporal drivers of nutrient dynamics in the Delta.
\acresetall

@@

Nutrient time series of monthly observations from 1976 to 2013 were obtained for ten active sampling stations in the Delta \citep[\cref{fig:delt_map,tab:stdescrp},][]{IEP13}.  Stations were grouped by location in the study area for comparison: \textit{peripheral} Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, and P8; \textit{interior} Delta stations D19, D26, and D28; and \textit{Suisun} stations D4, D6, and D7.  These stations cover all of the major inflows and outflows to the Delta and were selected for analysis based on the continuity of the period of observation \citep{Jabusch09}. Although many other stations are available for the region, the stations were chosen because they represent a consistent long-term dataset generated by a single program and they capture dominant seasonal and annual modes of nitrogen variability characteristic of the region \citep{Jabusch16}. Time series were complete for all stations except for an approximate ten year gap from 1996-2004 for D19.  Data were minimally processed, with the exception of averaging replicates that occurred on the same day.  The three nitrogen forms that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two). Less than 3\% of all observations were below the detection limit (left-censored), although variation was observed between analytes and location.  The ammonium time series had the most censored observations at sites C10 (25.4\% of all observations), MD10 (18.1\%), D28 (17.8\%), D19 (12\%), and D7 (7.9\%).

\ac{wrtds} models require flow data paired with nutrient data. At the Delta stations, daily flow estimates were matched with the corresponding sample dates for the nutrient data. Daily flow estimates were obtained from the Dayflow software program \citep{IEP16}.  The Sacramento daily flow time series was used to account for flow effects at C3, D19, D26, D28, and MD10, and the San Joaquin time series was used for C10 and P8 based on station proximity to each inflow. Given the complexity of inflows and connectivity of the system, only the inflow estimates from the Sacramento and San Joaquin rivers were used as measures of freshwater influence at each station.  Initial analyses indicated that model fit was not significantly improved with flow estimates from locations closer to each station, nor was model fit improved using lagged times series. Salinity was used as a proxy for flow at the Suisun Bay sites D4, D6, and D7 where tidal influences were much stronger. Salinity has been used as a tracer of freshwater influence for the application of \ac{wrtds} models in tidal waters \citep{Beck15}.  Models were evaluated using salinity at the Delta stations, but performance was reduced relative to models that used daily flow estimates.

@@

The observed time series for the ten Delta - Suisun Bay stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).  \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.  However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D26) had higher concentrations of ammonium during winter months when assimilation rates are lower \citep{Novick15}.  By location, observed concentrations of \ac{din} for the entire time series were higher on average for the peripheral Delta stations (C3, C10, MD10, P8; mean $\pm$ s.e.: $1.04$$\pm$$0.03$ \mgl) and similar for the interior Delta (D19, D26, D28, $0.43$$\pm$$0.01$) and Suisun Bay stations (D4, D6, D7, $0.44$$\pm$$0.01$).  Average concentrations were highest at P8 ($1.63$$\pm$$0.05$ \mgl) and lowest at C3 ($0.4$$\pm$$0.01$) for \ac{din}, highest at P8 ($0.28$$\pm$$0.02$) and lowest at D28 ($0.05$$\pm$$0.003$) for ammonium, and highest at C10 ($1.4$$\pm$$0.04$) and lowest at C3 ($0.15$$\pm$$0.004$) for nitrite/nitrate. Mean observed concentrations were also higher later in the time series for all analytes.  For example, average \ac{din} across all stations was $0.61$$\pm$$0.01$ \mgl for 1976-1995, compared to $0.7$$\pm$$0.01$ for 1996-2013. Seasonal changes across all years showed that nitrogen concentrations were generally lower in the summer and higher in the winter, although observed patterns were inconsistent between sites.  For example, site MD10 had distinct seasonal spikes for elevated \ac{din} in the winter, whereas other stations had less prominent seasonal maxima (e.g., C3, D7, \cref{fig:obsdat}).

Relative to the observed data, long-term trends between stations for the different nitrogen analytes were apparent from the modelled results (\cref{fig:prddat}).  Although each station varied in the overall concentrations, patterns within the three regions (peripheral Delta, interior Delta, and Suisun Bay) were observed.  Concentrations for all nitrogen analytes were highest in the peripheral stations.  Ammonium concentrations were highest at P8 and C3 and showed a consistent increase over time, followed by a reduction beginning in the early 2000s, whereas ammonium concentrations at C10 and MD10 were low and gradually decreasing throughout the period of record.  By contrast, \ac{din} and nitrite/nitrate concentrations at the peripheral Delta stations showed increases at  P8 and C10 followed by a decline in the early 2000s, whereas concentrations at C3 and MD10 were lower and did not show any noticeable trends.  Trends at the interior Delta stations showed a gradual increase in ammonium followed by a gradual decrease beginning in the early 1990s, particularly for D26.  Trends in \ac{din} and nitrite/nitrate for the interior Delta stations showed a reduction early in the time series, followed by a slight increase beginning in the mid-1980s, and finally a reduction beginning in the late 1990s.  These trends were similar for the Suisun Bay stations, although the reduction in the late 1990s did not occur.  By contrast, ammonium concentrations were low in Suisun Bay but a gradual increase over the period of record was observed.

@@

Water quality conditions in the Delta-Suisun Bay region are dynamic and not easily characterized from observed time series.  Annual aggregations of \ac{wrtds} modelled results and application of formal trend analyses provided insight into the spatial and temporal variation of nitrogen analytes in three distinct regions that is not possible with raw observations.  A general conclusion is that nitrogen concentrations have showed a consistent decrease beginning in the mid-1990s and early 2000s, although average concentrations remain above those observed earlier in the period of record.  These results are confirmed visually from \ac{wrtds} (\cref{fig:prddat}) and through significant results from trend tests (\cref{fig:trndcomp1}, \cref{tab:trndsann}).  Although the overall trends suggest a system-wide reduction, considerable differences by location and analyte were characterized by the analysis and are important independent of overall trends.  Nutrient concentrations were highest at peripheral Delta stations (C3, MD10, P8, and C10) that monitor inflows from the Sacramento and San Joaquin rivers.  The highest concentrations among all nitrogen analytes, ammonium in particular, were observed at P8 in the early 2000s as a direct consequence of \ac{wwtp} inputs upstream prior to infrastructure upgrades.  Elevated ammonium concentrations were also observed at C3 as a measure of upstream contributions from the Sacramento \ac{wwtp}.  By contrast, nitrite/nitrate concentrations were highest at C10 as a measure of contributions from the San Joaquin River to the south that drains a predominantly agricultural watershed.

Differing magnitudes of nitrogen analytes between stations as a function of source type can have an effect on the relationship between flow and nutrients.  Both \citet{Hirsch10} and \citet{Beck15} used \ac{wrtds} results to demonstrate variation between flow and nutrient dynamics depending on pollutant sources.  In particular, a chemodynamic (i.e., changing) response of nutrients with flow variation is common if nutrients originate primarily from the watershed through diffuse sources \citep{Thompson11,Wan17}.  Increased flow may induce a change in nutrient concentrations, such that reduction may occur with flushing or an increase may occur through mobilization.  By contrast, nutrient loads are relatively chemostatic or invariant with changes in flow if point-sources are the dominant contributor.  These relationships are modelled particularly well with \ac{wrtds}, which can provide a means of hypothesizing unknown sources or verifying trends in response to management actions.  As noted above, C10 at the inflow of the San Joaquin River is dominated by nitrite/nitrate consistent with diffuse, agricultural inputs from the watershed.  A logical expectation is that trends from observed data may vary considerably from trends with modelled results that are flow-normalized.  Accordingly, trend analysis of nitrate/nitrate by year and season showed that percent changes at C10 were typically underestimated with the observed data during the recent period from 1996-2013 (\cref{tab:trndsann,tab:trndsaft}).  This is consistent with an expected effect of flow on raw time series, particularly for chemodynamic behavior at locations that drain highly altered watersheds \citep{Wan17}.

% obs v flow norm trends
Our results underscore the importance of considering flow effects in the evaluation of water quality trends. There were important differences in trends between observed and flow-normalized data. These differences ranged from a simple change in the magnitude and significance of a trend, to more problematic changes where the flow-normalized trend could demonstrate a complete reversal relative to the observed (e.g., \ac{din} trends for all Suisun stations from 1996-2013, \cref{fig:trndcomp1}). Differences in apparent trends underscore the importance of considering flow effects in the interpretation of environmental changes, particularly if trend evaluation is used to assess the effects of nutrients on ecosystem health or the effectiveness of past nutrient management actions. An alternative evaluation of flow in the Delta demonstrated that flow contributions from different sources vary considerably over time at each station \citep{Novick15}.  For example, flow at MD10 represents seasonally changing inputs from the Sacramento and San Joaquin rivers and other sources. For simplicity, our analysis considered only the dominant water source and used the total daily inflow estimates from either the Sacramento River or the San Joaquin River at any of the Delta sites. Given that substantial differences with flow-normalized results were apparent from relatively coarse estimates of flow contributions, more precise differences could be obtained by considering the influence of multiple flow components at each location. Output from the Dayflow software program \citep{IEP16} provides a complete mass balance of flow in the Delta that could be used to develop a more comprehensive description.  Our analysis is the first attempt to model nutrient dynamics related to flow in the entire Delta, such that additional work should focus on improving the characterization of flow signals at each station.

Long-term trends in nutrient and phytoplankton concentations in Suisun Bay have also been the focus of intense study for many years \citep{Cloern83,Lehman92,Dugdale07,Jassby08,Glibert14}. Our results demonstrated an overall increase in ammonium from the late 1970s to 2000, with initial flow-normalized values of annual averages estimated as approximately 0.05 mg L$^-1$ for the Suisun stations in 1976 to a maximum ranging from 0.08 mg L$^{-1}$ (station D4) to above 0.1 mg L$^{-1}$ (D6) in 2000 (\cref{fig:prddat}).  Trends from 1996-2005 evaluated in \citep{Jassby08} showed a similar increase in ammonium.  However, a reversal of trends in recent years may also be occuring in Suisun Bay, as model estimates suggest either relatively constant concentrations or even a decrease at some stations beginning around 2000 (e.g., D6, D7, \cref{fig:trndcomp1}).  Combined with the shift towards a dominant spring peak in chloropyll growth (\cref{fig:clmchl}f), changing nitrogen ratios continue to be a concern for the management of production in the upper \ac{sfe}.

@@

The results for Suisun Bay provide additional descriptions of change in production as it relates to flow, grazing, and nitrogen ratios.  In general, clam biomass was associated with a decrease in \ac{chla} concentration, as shown by others \citep{Alpine92,Thompson08}. Our results also suggested that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima. This supports past research that showed a decrease in silica uptake by diatoms following invasion \citep{Cloern96,Kimmerer05}. A nontrivial portion of the \ac{din} increase could be related to the decrease in a major `sink', i.e., decreased \ac{din} uptake by phytoplankton due to top down grazing pressure from \textit{P. amurensis}.  Flow effects on phytoplankton production have also changed over time. In the absence of benthic grazing prior to invasion, \ac{chla} production was limited at low flow as less nutrients were exported from the Delta, stimulated as flow increases, and reduced at high flow as either nutrients or phytoplankton biomass are exported to the estuary (\cref{fig:clmchl}j). Recent years have shown a decrease in overall chlorophyll, with particularly low concentrations at low flow (high salinity).  As such, \ac{chla} production in early years is directly related to flow, whereas the relationship with flow in later years is indirect as increased flow reduces clam abundance and releases phytoplankton from benthic grazing pressure. These relationships have been suggested by others \citep{Cloern83,Alpine92,Parchaso02,Jassby08}, although the precise mechanisms demonstrated by \ac{wrtds} provide additonal quantitative evidence of factors that drive water quality in the Delta.

@@

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth,page=1]{figs/delt_map.pdf}
\caption{The \acl{sfe} and Delta region with monitoring stations used for analysis. The Delta drains the combined watersheds of the Sacramento and San Joaquin rivers (inset). The grey outline is the legal boundary of the Delta. All data were obtained from the \acl{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}, \cite{IEP13}). See \cref{tab:stdescrp} for station descriptions.}
\label{fig:delt_map}   
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth,page=1]{figs/trndcomp1.pdf}
\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes. Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$). Trends are shown separately for different annual groupings from 1976 to 1995 (top) and 1996 to 2013 (bottom). See \cref{fig:trndcomp2,fig:trndcomp3} for seasonal groupings.}
\label{fig:trndcomp1}   
\end{figure}

@@

\caption[Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County]{Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County.  Wastewater discharge requirements were implemented in 2006 to convert ammonium to nitrate.}\label{fig:stock}
\end{figure}
